Score-based tests of measurement invariance: Use in practice

### 2014-03-14 frontiers/IMPSpres.Rnw
update presentation. almost finished. need to practice orally.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@329 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{frame}
\frametitle{Problem Scenario}

\end{frame}
%--------------------------------------------------------------------------

\begin{frame}
\frametitle{Overview} 
% Table of contents slide, comment this block out to remove it
\tableofcontents 
% Throughout your presentation, if you choose to use \section{} and 
%\subsection{} commands, these will automatically be printed on this slide 
%as an overview of your presentation
\end{frame}
%--------------------------------------------------------------------------
\section{Theoretical Background}
\section{Simulation}
\subsection{Violation Parameter is IN the Model}
\begin{frame}
 \begin{figure}
    \caption{Model used for the simulation.}
    \label{fig:pathinmodel}
\resizebox{8.0cm}{!}{
\begin{tikzpicture}
[latent/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=20mm, minimum height=10mm},
 manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
                inner sep=0pt,minimum width=20mm, minimum height=10mm},
 error/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=10mm},
 post/.style={->,shorten >=1pt,>=stealth',semithick},
 postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
 postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
 sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=20mm,semithick},
 errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=10mm,semithick}]

@@

\node[error] (E1) at (7,6) {E1}
   edge [post] node[auto,swap]{\small{1}}(SCALE1)
   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{11}=26.77$}} (E1.45);
\node[error] (E2) at (7,4.5) {E2}
   edge [post] node[auto,swap]{\small{1}} (SCALE2)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{22}=13.01$}}(E2.45);
\node[error] (E3) at (7,3) {E3}
   edge [post] node[auto,swap]{\small{1}}(SCALE3)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{33}=30.93$}}(E3.45);
\node[error] (E4) at (7,1) {E4}
   edge [post] node[auto,swap]{\small{1}}(SCALE4)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{44}=3.17$}}(E4.45);
\node[error] (E5) at (7,-0.5) {E5}
   edge [post] node[auto,swap]{\small{1}}(SCALE5)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{55}=8.82$}} (E5.45);
\node[error] (E6) at (7,-2) {E6}
   edge [post] node[auto,swap]{\small{1}}(SCALE6)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{66}=22.5$}}(E6.45);
\end{tikzpicture}}
\end{figure}
\end{frame}

\begin{frame}
\begin{figure}
\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
    parameters are $\lambda_{11}$ and $\lambda_{11} \ldots \lambda_{62}$.}}
\label{fig:sim11res}
<<sim11res, fig=TRUE, height=5, width=9, echo=FALSE>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
       subset = (parms == "loading" & 
          pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
       type = "b", xlab = 
          expression(paste("Violation Magnitude (", lambda[11], ")")),
       ylab = "Power", key = mykey, as.table = TRUE,
       strip = function(..., which.given, factor.levels){
           if(which.given == 2){
             strip.default(which.given, factor.levels = 
                           parlabs[c(1, 5)], ...)
	   } else {
	       strip.default(which.given, factor.levels = factor.levels, ...)
	   }
       })
@

\end{figure}
\end{frame}
%--------------------------------------------------------------------------
\begin{frame}
\begin{figure}
\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
    parmaeter is $\psi_{11} \ldots \psi_{66}, \sigma_{12}$ and
    $\mu_{11} \ldots \mu_{66}$.}}
\label{fig:sim12res}
<<sim12res, fig=TRUE, height=5, width=9, echo=FALSE>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
       subset = (parms == "loading" & 
          pars %in% c("All Errors", "Covariance", "All Intercepts") 
                 & diff %% 0.5 == 0),
       type = "b", xlab = 
          expression(paste("Violation Magnitude (", lambda[11], ")")),
       ylab = "Power", ylim = c(0,1), key = mykey, as.table = TRUE,
       strip = function(..., which.given, factor.levels){
           if(which.given == 2){
             strip.default(which.given, factor.levels = 
                           parlabs[c(2, 6, 7)], ...)
	   } else {
	       strip.default(which.given, factor.levels = factor.levels, ...)
	   }
       })
@

@@

\subsection{Violation Parameter is OUT the Model}
\begin{frame}
\begin{figure}
    \caption{Model used for the simulation.}
    \label{fig:pathinmodel}
\resizebox{8.0cm}{!}{
\begin{tikzpicture}
[latent/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=20mm, minimum height=10mm},
 manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
                inner sep=0pt,minimum width=20mm, minimum height=10mm},
 error/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=10mm},
 post/.style={->,shorten >=1pt,>=stealth',semithick},
 postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
 postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
 sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=20mm,semithick},
 errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=10mm,semithick}]

>>>>

\begin{frame}
\frametitle{Problem Scenario}
\begin{itemize}
  \item
 \emph{Data:} \\
 The data consist of responses to a gratitude scale from $n=1401$ youth 
 aged 10--19 years, dividing into six age groups: 10-11, 12-13, 14,
 15, 16, 17-19. 
  \item
\emph{Question:} \\
The researcher is interested in whether the scale
is measurement invariant across these age groups.  
\end{itemize}

\end{frame}
%--------------------------------------------------------------------------
\begin{frame}
  \frametitle{Similar Problem Scenarios and Generalization}
  \begin{itemize}
    \item Whether model parameter(s) hold(s) for people with different
      income level, grade level, motivation level, etc.  
    \item In general, the problem is whether measurement invariance(M.I.)
      hypothesis holds with respect to an auxiliary ordinal variable. 
    \end{itemize}
\end{frame}
%--------------------------------------------------------------------------
\begin{frame}
\frametitle{Possible Solution}
\begin{itemize}
\item Likelihood Ratio Test(LRT) for full model and reduced model(s).
  \begin{itemize}
    \item Ignore the nature of ordinal variable, treat it as
      categorical varaible. 
    \item Sensitive to sample size. Rejecte the reduced model(s) too often.
    \item Rather tedious and time consuming to fit and compare all possible
      models. e.g, for testing one parameter, you need to fit $2^{m-1}$ 
      models, $m$ is the level of ordinal variable.
    \end{itemize}
\item Score-based Tests w.r.t. ordinal variable. 
\end{itemize}
  \end{frame}
\begin{frame}
\frametitle{Overview} 
% Table of contents slide, comment this block out to remove it
\tableofcontents 
% Throughout your presentation, if you choose to use \section{} and 
%\subsection{} commands, these will automatically be printed on this slide 
%as an overview of your presentation
\end{frame}
%--------------------------------------------------------------------------
\section{Theoretical Background}
\begin{frame}
  \frametitle{Score for Each Observation}
\begin{equation}
\label{eq:score}
s(\theta;x_{i}) = \left (
  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{1}}, \dots,
  \frac{\partial \ell(\theta;x_{i})}{\partial \theta_{k}} \right
  ),\ i=1,\ldots,n,
\end{equation}

\vspace{0.3in}
where $\ell(\theta; x_i)$ is the likelihood associated with individual
$i$ and $\theta$ is a $k$-dimensional parameter vector. The corresponding
maximum likelihood estimate $\hat{\theta}$ solves the first order condition:
$\sum_{i = 1}^n s(\hat \theta;x_{i}) = 0$. 
\end{frame}
%--------------------------------------------------------------------------

\begin{frame}
  \frametitle{Cumulative Sum Score: Decorrelate and Order}
\begin{equation} 
    \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}

\vspace{0.3in}
where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
nt \rfloor$ is the integer part of $nt$, $t$ is the cumulative proportion 
of $n$, and $x_{(i)}$ reflects the 
individual with the $i$-th smallest value of the auxiliary variable $V$. 
  \end{frame}
%--------------------------------------------------------------------------
\begin{frame}
  \frametitle{Convergence Distribution}
  \begin{itemize}
   \item  Under the hypothesis of measurement invariance, a central limit 
     theorem can be
   used to show that the fluctuation of the above cumulative sum score
   matrix ${\bm B}(\cdot; \hat {\bm \theta})$ converge in distribution to a 
   k-dimensional Brownian bridge ${\bm B}^{0}(\cdot)$.
   \begin{equation}
     \label{eq:converge}
    {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}
        {\rightarrow}~ {\bm B}^{0}(\cdot),
   \end{equation}
   \item This result allows us to calculate $p$-values and
   critical values for test statistics under the hypothesis of
   measurement invariance.
  \end{itemize}
 \end{frame}
%--------------------------------------------------------------------------  
 \begin{frame}
   \frametitle{Statistics Based on Cumulative Sum Socre}
   Multiple test statistics are available, depending on how one
summarizes the behavior of the cumulative sum of scores. Below shows
 two statistics for ordinal auxiliary variable with $m$ levels,

\scriptsize{\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
        \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
          \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2
\end{eqnarray}}
where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$,
$(t_{ell}$ is the cumulative proportions associated with level $\ell$, 
($\ell=1,\ldots,m-1)$.
   \end{frame}
%-------------------------------------------------------------------------- 
\begin{frame}
\frametitle{What If the Auxiliary Variable is Indeed Categorical}
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
    \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}
      (\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}

\vspace{0.3in}
This statistic is asymptotically equivalent to the likelihood
ratio test statistic but only requires estimation of only one model 
(the reduced model).
 \end{frame}
%--------------------------------------------------------------------------
\begin{frame}
  \frametitle{Visual Explanation of the Score-based Test w.r.t. ordinal
    auxiliary variable.}
$$\bordermatrix{\text{}& \theta_{1} & \theta_{2} & \ldots &\theta_{k}\cr
                x_{1}&  1.32  & -4.32 & \ldots & -2.34\cr
                x_{2}&  3.37 & 8.32 & \ldots & \textcolor{red}{63.45}\cr
                x_{..}& \vdots & \vdots & \vdots & \vdots\cr
                x_{m-1}&  12.34 & 13.32 & \ldots & 6.89\cr
                x_{m}& 0     &   0     & \ldots & 0} $$

\section{Simulation}
\subsection{M.I. Violation Parameter is IN the Model}
\begin{frame}
 \begin{figure}
    \caption{Model used for the simulation.}
    \label{fig:pathinmodel}
\resizebox{8.0cm}{!}{
\begin{tikzpicture}
[latent/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=20mm, minimum height=10mm},
 manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
                inner sep=0pt,minimum width=20mm, minimum height=10mm},
 error/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=10mm},
 post/.style={->,shorten >=1pt,>=stealth',semithick},
 postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
 postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
 sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=20mm,semithick},
 errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=10mm,semithick}]

@@

\node[error] (E1) at (7,6) {E1}
   edge [post] node[auto,swap]{\small{1}}(SCALE1)
   edge [errorsling, bend right=100]node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{11}=26.77$}} (E1.45);
\node[error] (E2) at (7,4.5) {E2}
   edge [post] node[auto,swap]{\small{1}} (SCALE2)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{22}=13.01$}}(E2.45);
\node[error] (E3) at (7,3) {E3}
   edge [post] node[auto,swap]{\small{1}}(SCALE3)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{33}=30.93$}}(E3.45);
\node[error] (E4) at (7,1) {E4}
   edge [post] node[auto,swap]{\small{1}}(SCALE4)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{44}=3.17$}}(E4.45);
\node[error] (E5) at (7,-0.5) {E5}
   edge [post] node[auto,swap]{\small{1}}(SCALE5)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{55}=8.82$}} (E5.45);
\node[error] (E6) at (7,-2) {E6}
   edge [post] node[auto,swap]{\small{1}}(SCALE6)
   edge [errorsling, bend right=100] node[right=0.1,text width=2cm]
   {\scriptsize{$\psi_{66}=22.5$}}(E6.45);
\end{tikzpicture}}
\end{figure}
\end{frame}
%--------------------------------------------------------------------------
\begin{frame}
  \frametitle{Simulation Setup}
  Power and Type I error were examined across the following conditions.
 \begin{itemize}
   \item Three sample sizes ($n=120,\textcolor{red}{480},960$), 
   \item Three numbers of categories ($m=4,8,12$), 
   \item Violation magnitude: $d$ times the parameters' asymptotic 
     standard errors (scaled by $\sqrt n$), with $d=0,0.25,0.5,...,4$ 
   \item Violation Level: began at level $1+m/2$ of the auxiliary variable 
         $V$
 \end{itemize}
   Statistics $\mathit{WDM}_o$, $\max \mathit{LM}_o$, $\mathit{LM}_{uo}$ were 
   examined.
\end{frame}
%--------------------------------------------------------------------------
\begin{frame}
\begin{figure}
\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
    parameters are $\lambda_{11}$ and $\lambda_{11} \ldots \lambda_{62}$.}}
\label{fig:sim11res}
<<sim11res, fig=TRUE, height=5, width=9, echo=FALSE>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
       subset = (parms == "loading" & 
          pars %in% c("Loading1", "All Loadings") & diff %% 0.5 == 0),
       type = "b", xlab = 
          expression(paste("Violation Magnitude (", lambda[11], ")")),
       ylab = "Power", key = mykey, as.table = TRUE,
       strip = function(..., which.given, factor.levels){
           if(which.given == 2){
             strip.default(which.given, factor.levels = 
                           parlabs[c(1, 5)], ...)
	   } else {
	       strip.default(which.given, factor.levels = factor.levels, ...)
	   }
       })
@

\end{figure}
\end{frame}
%--------------------------------------------------------------------------
\begin{frame}
\begin{figure}
\tiny{\caption{M.I.violation parameter $\lambda_{11}$.  Testing
    parameters are $\psi_{11} \ldots \psi_{66}, \sigma_{12}$ and
    $\mu_{11} \ldots \mu_{66}$.}}
\label{fig:sim12res}
<<sim12res, fig=TRUE, height=5, width=9, echo=FALSE>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1$test), points = TRUE, lines = TRUE)
xyplot(power ~ diff | nlevels + pars, group = ~ test, data = sim1,
       subset = (parms == "loading" & 
          pars %in% c("All Errors", "Covariance", "All Intercepts") 
                 & diff %% 0.5 == 0),
       type = "b", xlab = 
          expression(paste("Violation Magnitude (", lambda[11], ")")),
       ylab = "Power", ylim = c(0,1), key = mykey, as.table = TRUE,
       strip = function(..., which.given, factor.levels){
           if(which.given == 2){
             strip.default(which.given, factor.levels = 
                           parlabs[c(2, 6, 7)], ...)
	   } else {
	       strip.default(which.given, factor.levels = factor.levels, ...)
	   }
       })
@

@@

\subsection{M.I. Violation Parameter is OUT the Model}
\begin{frame}
\begin{figure}
    \caption{Model used for the simulation.}
    \label{fig:pathinmodel}
\resizebox{8.0cm}{!}{
\begin{tikzpicture}
[latent/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=20mm, minimum height=10mm},
 manifest/.style={rectangle,draw=black!50,fill=white!20,thick,
                inner sep=0pt,minimum width=20mm, minimum height=10mm},
 error/.style={circle,draw=black!50,fill=white!20,thick,
                inner sep=0pt, minimum width=10mm},
 post/.style={->,shorten >=1pt,>=stealth',semithick},
 postred/.style={->,shorten >=1pt,>=stealth',semithick,red},
 postdashed/.style={->,shorten >=1pt,>=stealth',semithick,red,dashed},
 sling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=20mm,semithick},
 errorsling/.style={<->,shorten >=1pt,>=stealth',shorten <=1pt,>=stealth',
   auto,distance=10mm,semithick}]
